---
title: "AB005_prepro_MZMine"
output: html_document
date: "2024-05-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Script for preprocessing & selecting untar ions

## 0. Set-up

```{r}
library(tidyverse)
library(data.table)
```

## 1. Read-in data

```{r}

data_list = list.files("input_folder", pattern = "^MZMINE_.*_quant\\.csv$", full.names = TRUE)

data = lapply(data_list, fread)
#data = lapply(data, setDF)

# for loop throuhg list 
# run function
mzmine_peaks = list()
mzmine_areas = list()
for(df in data){
  
  # tidy data 
  df$feature_ID = paste0("FT", df$`row ID`)
  # separate the peak_info from the areas 
  mzmine_peak = df[, c(1:13)]
  mzmine_peak$feature_ID = df$feature_ID
  mzmine_area = df %>% 
    dplyr::select(contains("pool"))
  mzmine_area = cbind(df$feature_ID, mzmine_area)
  
  strain = sapply(strsplit(as.character(colnames(df)[26]), "_"), '[', 2)
  # add dfs to new lists 
  mzmine_peaks[[strain]] = mzmine_peak
  mzmine_areas[[strain]] = mzmine_area
}

lapply(mzmine_peaks, dim)

#mzmine_areas = mzmine_areas[-4]

```

## 2. Compare pools to find potential drug metabolites

```{r}
# read-in pool file
pools = read_csv(file.path("input_folder", "NEW_pooling_scheme_info_with_IS.csv"))
pools = pools %>% 
  dplyr::filter(!(grepl("IS_", Product.Name)))

# function to transpose and prep each df 
prep_df = function(complete_df){
  
  # transpose df 
  my_dataset_t = setDT(data.table::transpose(complete_df, keep.names = "rn"))
  
  # set colnames for my_dataset_t
  colnames(my_dataset_t) = as.character(c("sample_name", complete_df$V1))
  my_dataset_t = my_dataset_t[-1, ]
  
  # set abundances columns as numeric 
  my_dataset_t[, 2:dim(my_dataset_t)[2]] <- 
    lapply(my_dataset_t[,2:dim(my_dataset_t)[2]], as.numeric)
  
  # log-transform data 
  prelog_data = my_dataset_t[,2:dim(my_dataset_t)[2]]
  # set 0 values to 1 
  prelog_data[prelog_data == 0] <- 1
  log_data <- log(prelog_data, 2)
  
  # include pool & TP metadata 
  sample_name = my_dataset_t$sample_name
  log_data = cbind(sample_name, log_data)
  log_data$pool = sapply(strsplit(as.character(log_data$sample_name), "_"), '[', 3)
  log_data$TP = sapply(strsplit(as.character(log_data$sample_name), "_"), '[', 4)
  
  log_data = log_data %>% 
    dplyr::filter(TP == "TP8")
  
}

# run function for mzmine_areas 
log_mzmine_areas = lapply(mzmine_areas, function(x){prep_df(complete_df = x)})



# function to compare all TPs of pools with and without drug
compare_drug = function(log_data, pool_file){ 
  
  # compare per drug 
  # compare all metabolites in drug-containing pools (n=4) vs all other pools (n=28)
  
  # read-in drug-pools 
  pool_info = pool_file
  pool_info = na.omit(pool_info)
  
  # make list of pools per drug 
  to_include = apply(pool_info, 1, function(x) colnames(pool_info)[which(x == 1)])
  names(to_include) = pool_info$Product.Name
  
  # set empty list to collect wilcoxon results 
  results = list()
  
  for (d in names(to_include)){
    
    # select data per drug 
    df1 = log_data %>%
      filter(pool %in% to_include[[d]]) %>% 
      dplyr::select(-c(sample_name, pool, TP)) %>% 
      as.matrix() 
    
    df2 = log_data %>%  
      filter(!pool %in% to_include[[d]]) %>% 
      dplyr::select(-c(sample_name, pool, TP)) %>% 
      as.matrix()
    
    # Create an empty matrix
    pvalue = matrix(NA, ncol(df1), ncol=4)
    
    # run T-test per drug & calculate fold change 
    for(i in 1:length(colnames(df1))){
      FC = mean(df1[,i], na.rm = T) - mean(df2[,i], na.rm=T) # FC is calculated by subtraction, because values are already log2-transformed
      test_result = t.test(df1[,i], df2[,i], alternative = "two.sided", paired=F)
      pvalue[i, 1] <- test_result$p.value
      pvalue[i, 2] <- test_result$statistic
      pvalue[i, 4] <- FC
      row.names(pvalue) <- colnames(df1)
    }
    
    # do p-value adjustment
    pvalue[, 3] <- stats::p.adjust(pvalue[, 1], method="BH",
                                   n=length(na.omit(pvalue[, 1])))
    
    colnames(pvalue) = c("pvalue", "statistic", "pajd", "FoldChange")
    
    results[[d]] = pvalue
    
  } 
  
  return(results)

}


# run function for mzmine_areas 
TP8_pools_results = lapply(log_mzmine_areas, function(x){compare_drug(log_data = x, pool_file = pools)})

#saveRDS(TP8_pools_results, "TP8_ttest_results.RDS")
#TP8_pools_results_save = TP8_pools_results
#save(pools_results, file ="AB005_mzmine_pools_results.RData")
#load("AB005_mzmine_pools_results.RData")

TP8_pools_results = readRDS(file.path("input_folder", "TP8_ttest_results.RDS"))

# add rownames as column 
pools_results = lapply(save_pools_results, function(x){
  for(i in x){
  i=i %>% as.data.frame(i) %>% 
    rownames_to_column(var = "feature_ID")}
  return(x)})


# run for loop to display pvalue hists 
p_list = list()
n=0
for(i in TP8_pools_results){
  
  p_vals = lapply(i, function(x) x[x[,3]]) # take adj p-val
  p_vals = unlist(p_vals)
  #p_vals = do.call(p_vals, function(x) unlist(x))
  p_vals = reshape2::melt(p_vals)
  
  n=n+1
  p_list[[n]] = ggplot(p_vals, aes(x=value))+
    geom_histogram()
  
}
library(gridExtra)
do.call("grid.arrange", c(p_list))


#### select only the ions for which p-val < 0.01 ####
select_ions = function(results){ 
  library(purrr)
  results = lapply(results, function(x) as.data.frame(x))
  selected_ion = purrr::map(results, ~.x %>% filter(pajd < 0.01 & abs(FoldChange) > 2))
  
  test = bind_rows(selected_ion, .id = "column_label")
  colnames(test) <- c("drug", "pval", "statistic", "padj", "FoldChange")
  
  
  ions = lapply(selected_ion, function(x) rownames(x))
  # gather ions 
  ions = ions %>% 
    purrr::map(as_tibble) %>% 
    bind_rows(., .id = "column_label")
  
  # add ion names to test
  ions = cbind(test, ions) 
  #ions$value = as.character(ions$value)
  
  
  return(ions)
}


results = lapply(TP8_pools_results, select_ions)

```

## 2B. Match ions to ion_info

```{r}

names(results)
names(mzmine_peaks)
pool_info = na.omit(pools)

ions_info = list()
for (i in seq_along(results)) {
  
  strain = names(results)[i]
  df = results[[i]]
  
  peaks = mzmine_peaks[[strain]]
  
  # get peak info on ions
  ions = left_join(df,
                   peaks[,c(2,3,9,13,14)],
                   by = c("value" = "feature_ID"))
  
  
  ions = left_join(ions, 
                        pool_info[30:32], 
                        by = c("drug" = "Product.Name"))
  
  ions_info[[strain]] = ions 
  
  
  
}

# bind ions from different experiments together in one df 
results = bind_rows(ions_info, .id = "strain")
table(results$strain)
length(unique(results$drug))

```

##3A. Filter for different mass and RT than parent drug

```{r}

# calculate deltaRT between parent and potential metabolite 
my_df = left_join(results, pool_info[, c(30, 35)], 
                  by = c("column_label" = "Product.Name"))

my_df$deltaRT = my_df$RT.NR - my_df$`row retention time`

# calculate deltaMZ between parent and potential metabolite 
my_df$deltaMZ = my_df$`row m/z` - my_df$M.SD

# check distribution of delta values 
hist(my_df$deltaMZ)
hist(my_df$deltaRT)


# make metabolite selection: different m/z than parent compound & different RT 
my_df_fil = my_df %>% 
  filter(abs(deltaMZ) > 1 & (deltaRT < -0.3 | deltaRT > 0.3))
nrow(my_df)
nrow(my_df_fil)

```

##3B. Filter for drug-specificity

```{r}
# round up mz value for easier matching
my_df_fil$round_mz = round(my_df_fil$`row m/z`, digits = 2)
drug_candidates = my_df_fil %>% 
  group_by(strain, round_mz) %>% 
  mutate(ion_count = n_distinct(drug)) 

hist(drug_candidates$ion_count)
ggplot(drug_candidates, aes(x=ion_count)) + 
  geom_histogram() + 
  facet_wrap(~strain)

drug_candidates = drug_candidates %>%
  dplyr::filter(ion_count == 1)
nrow(drug_candidates)

drug_candidates %>% 
  group_by(strain) %>% 
  summarise(ion_count = n_distinct(value))

```

##3C. Delete ions that are also for that drug in nobug

```{r}
# test whether mz value of ion overlaps with nobug experiment 
test = drug_candidates %>% 
  group_by(round_mz) %>% # check mz value & RT match up until 2 sec /2 decimals 
  summarise(shared = paste(strain, collapse = ",")) %>% 
  mutate(exclude = str_detect(shared, "nobug"))

test = test[test$exclude == TRUE, ]

# when mzvalue also in nobug experiment, delete ion from candidates 
candidates = anti_join(drug_candidates, test, 
                       by = "round_mz")

# check distribution 
hist(candidates$deltaMZ)
hist(candidates$deltaRT)

table(candidates$strain)


```


## 4A. Read all peaks files 
```{r}

# transpose dfs in allpeaks
allpeaks_t = lapply(mzmine_areas, function(x){ 
  setDT(data.table::transpose(x, keep.names = "rn"))})

# define colnames 
columns = lapply(allpeaks_t, function(x) as.character(x[1, ]))
# delete first row 
allpeaks_t = lapply(allpeaks_t, function(x) dplyr::slice(x, -1))

# set colnames to each df 
for (i in seq_along(allpeaks_t)) {
  
  colnames(allpeaks_t[[i]]) <- columns[[i]]
  
}

# set all columns to numeric 
allpeaks_t = lapply(allpeaks_t, function(x) 
  mutate_at(x, .vars = 2:dim(x)[2], .funs = as.numeric))

# set df for each data frame
allpeaks_t = lapply(allpeaks_t, setDF)


# define function to add metadata to each dataframe 
add_metadata = function(x){ 
  
  x$strain = sapply(strsplit(as.character(x[,1]), "_"), '[', 2)
  x$pool = sapply(strsplit(as.character(x[,1]), "_"), '[', 3)
  x$TP = sapply(strsplit(as.character(x[,1]), "_"), '[', 4)
  x$well_location = sapply(strsplit(as.character(x[,1]), "_"), '[', 5)
  x$date = sapply(strsplit(as.character(x[,1]), "_"), '[', 6)
  
  
  return(x)
}

allpeaks_t = lapply(allpeaks_t, add_metadata)
```

##4B. Calculate pool specificity 
```{r}
#### calculate fold change between TP0 and TP8 ####
library(data.table)
FCtime = data.frame(matrix(ncol=8,nrow=0, dimnames=list(NULL, c("FC_time", "value", "drug", "meanTP8", 
                                                                "mean_non", "median_non", "medianTP8", "strain"))))
# calculate FC of ion over time (0-8hours)
for (i in 1:nrow(candidates)) { 
  
  # collect info on ion
  value = candidates$value[i]
  strain = candidates$strain[i]
  drug = candidates$drug[i]
  
  df = allpeaks_t[[strain]]
  setDF(df)
  
  pools = to_include[[drug]]
  
  df$pool_include = df$pool %in% pools
  
  df_select = df %>% 
    dplyr::select(TP, !!value, pool_include) %>% 
    filter(pool_include == T) 
  
  TP0 = df_select %>% 
    filter(TP == "TP0")
  meanTP0 = mean(TP0[, 2], na.rm=T)
  
  
  TP8 = df_select %>% 
    filter(TP == "TP8")
  meanTP8 = mean(TP8[, 2], na.rm=T)
  medianTP8 = stats::median(TP8[, 2], na.rm=T)
  
  FC_time = meanTP8/meanTP0
  
  df_non = df %>% 
    dplyr::select(!!value, pool_include, TP) %>% 
    filter(pool_include == F) %>% 
    filter(TP == "TP8")
  
  mean_non = mean(df_non[,1], na.rm=T)
  median_non = median(df_non[,1], na.rm = T)
  
  # add FC_time to new dataframe 
  FCtime[i, 1] = FC_time
  FCtime[i, 2] = value
  FCtime[i, 3] = drug
  FCtime[i, 4] = meanTP8
  FCtime[i, 5] = mean_non
  FCtime[i, 6] = median_non
  FCtime[i, 7] = medianTP8
  FCtime[i, 8] = strain
  
  
}

# Check FCtime 
# check distribution of FCtime 
ggplot(FCtime, aes(x=log2(FC_time))) + 
  geom_density()


# add log2(FC) value 
FCtime$log2FCtime = log2(FCtime$FC_time)

# how many ions per strain? 
FCtime %>% 
  group_by(strain) %>% 
  summarise(count = n_distinct(value))


# select ions for which the area in the non-pools in low of the pools area at TP8
FCtime$pool_percentage = (FCtime$mean_non / FCtime$meanTP8)*100
FCtime$pool_median = (FCtime$median_non / FCtime$medianTP8)*100

ggplot(FCtime, aes(x=pool_percentage)) + 
  geom_density()
hist(FCtime$pool_percentage)

# select only if non detected in non-pools 
tmp = FCtime %>%
  #filter(FC_time > 1) %>%
  filter( pool_percentage <= 10 & pool_median <= 100)
hist(tmp$pool_percentage, breaks =60)
hist(tmp$pool_median, breaks =60)


### select candidates for FC ###
candidates = left_join(tmp, drug_candidates,
                                by = c("value", "strain", "drug"))

table(candidates$strain)

```


##4C. Test for drug-specificty
```{r}

names(allpeaks_t)
unique(candidates$strain)

# make list of pools per drug 
to_include = apply(pool_info, 1, function(x) colnames(pool_info)[which(x == 1)])
names(to_include) = pool_info$Product.Name

# set new empty dataframe 
candidate_df = data.frame(matrix(ncol=3,nrow=0, dimnames=list(NULL, c("value", "strain", "drug"))))

## loop over all candidate ions and check plot pools that contain the drug vs without
pdf("./Plots/MZmine_plots/candidates_timecourse_pools_241127.pdf")
for (i in 1:nrow(candidates)) { 
  
  # collect info on ion
  value = candidates$value[i]
  strain = candidates$strain[i]
  drug = candidates$drug[i]
  
  df = allpeaks_t[[strain]]
  
  pools = to_include[[drug]]
  
  df$pool_include = df$pool %in% pools
  
  df_temp = df %>% 
    filter(pool_include == T) %>% 
    dplyr::select(V1, strain, pool, TP, well_location, pool_include, all_of(value)) %>% 
    na.omit()
  
  include = length(unique(df_temp$pool))
    
  if(include >=3){ 
    candidate_df[i, 1] = value
    candidate_df[i, 2] = strain
    candidate_df[i, 3] = drug
  }
    
  if (include >= 3){
  p=ggplot(df, aes_string(x="TP", y=value, color="pool_include", label="pool")) +
    geom_point() +
    geom_line(aes(group=pool)) +
    geom_text(check_overlap = T) +
    labs(title= paste0("Timecourse ", value, " from: ", strain, " and", " drug: ", drug),
        x = "Time (hours)",
        y = "Raw peak area") +
    theme_minimal()

  print(p)
  }
  
}
dev.off()

candidate_df = na.omit(candidate_df)
table(candidate_df$strain)

# merge candidate df with candidates for info on ions 
integration_df = left_join(candidate_df, drug_candidates, 
                         by = c("drug",
                                "value", "strain"))

# save only relevant info 
integration_df = integration_df %>% 
  dplyr::select(value, strain, drug, padj, FoldChange, `row m/z`, `row retention time`, deltaMZ, deltaRT, M.SD, RT.NR)

table(integration_df$strain)

# write integration_df
write.csv(integration_df, "./240607_mzmine_candidates.csv")

```

## 4D. Subset integration_df for degraded drugs 
```{r}
degraded_drugs = read.csv(file.path("input_folder", "AB005_results_p02_10percent_241126.csv"))

setdiff(unique(degraded_drugs$strain), unique(integration_df$strain))

degraded_drugs$combo = paste0(degraded_drugs$Product.Name, "_", degraded_drugs$strain)
integration_df$combo = paste0(integration_df$drug, "_", integration_df$strain)

drug_metabolites = integration_df %>% 
  filter(combo %in% degraded_drugs$combo)

physio_metabolites = integration_df %>% 
  filter(!(combo %in% degraded_drugs$combo))

```


### 4D1. Plots for drug mets 
```{r}
# Function to standardize ggplot theme
cell_plot_theme <- theme(
  text = element_text(family = "Arial", size = 8), # Set Arial font and text size
  axis.text = element_text(size = 8), # Adjust axis text size
  axis.title = element_text(size = 10), # Adjust axis title size
  legend.text = element_text(size = 8), # Adjust legend text size
  legend.title = element_text(size = 8), # Adjust legend title size
  plot.title = element_text(size = 12, hjust = 0.5), # Center title
  panel.background = element_blank(), # Remove background color
  panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5), # Add border
  panel.grid.major = element_blank(), # Subtle grid
  panel.grid.minor = element_blank(), # No minor grid
  plot.margin = margin(5, 5, 5, 5) # Adjust margins
)

unique(drug_metabolites$strain)
# recode strain names 
drug_metabolites$strain <- recode(
  drug_metabolites$strain,
  "Anaeslundi" = "A. naeslundi",
  "Aomnicolens" = "A. omnicolens",
  "Btheta" = "B. thetaiotaomicron",
  "Buni" = "B. uniformis",
  "Ccomes" = "C. comes",
  "Cramosum" = "C. ramosum",
  "Cscindens" = "C. scindens",
  "Dformi" = "D. formicigenerans",
  "Ef2" = "E. faecalis",
  "Ghaemolysans" = "G. haemolysans"
)


# make barplot of mets per strain 
p = drug_metabolites %>% 
  group_by(strain) %>% 
  summarise(count = n_distinct(value)) %>% 
  ggplot(., aes(x=reorder(strain, -count), y=count)) +
  geom_bar(stat = "identity") + 
  coord_flip() +
  theme_minimal() +
  labs(y = "No. of drug metabolites") + 
  theme(axis.text.y = element_text(face = "italic"), 
        axis.title.y = element_blank()) +
  cell_plot_theme 

print(p)
ggsave("./Plots/AB005_untar_count.tiff", plot = p, dpi = 300, width = 3, height = 2, units = "in")

# plot distribution of mzvalues & deltaRT 
# make long df with delta values 
delta_long = pivot_longer(drug_metabolites, 
                          cols = c(deltaMZ, deltaRT),
                          values_to = "drug_info")

# plot delta RT distribution
delta_long %>% 
  filter(name == "deltaRT") %>%
  ggplot(., aes(y=drug_info, x=name, fill = name)) +
  geom_boxplot(width = .2, outlier.shape = NA) + 
  geom_jitter(width = .05, alpha = .3) +
  theme_minimal() +
  scale_fill_manual(values = c("deltaMZ" = "lightpink", "deltaRT" = "purple"))


# calculate deltamz prevalence 
delta_long$round_mz = round(delta_long$drug_info, digits = 1)
  
delta_long %>% 
  filter(name == "deltaMZ") %>% 
  group_by(round_mz) %>% 
  summarise(count = n_distinct(drug_info)) %>%
  arrange(desc(count)) %>% 
  ggplot(., aes(x=round_mz)) +
  geom_histogram() +
  cell_plot_theme + 
  labs(x="delta m/z value", y="")
ggsave("./Plots/AB005_deltaMZ.tiff", dpi = 600, width = 3, height = 2, units = "in")

```

